---
title: Contributing
output: github_document
---

```{r setup, echo=FALSE}
# install_github("hadley/emo")
grin <- emo::ji("grin")
mark_yes <- emo::ji("white_check_mark")
mark_no <- emo::ji("negative_squared_cross_mark")
```

## Package scope

The purpose of this package is to provide you the means:

- to build Vega-Lite chart-specifications.
- to render chart-specifications into HTML.
- to communicate your charts.

## Style

This package aspires to use the [Tidyverse Style Guide](http://style.tidyverse.org), with some minor modifications.

- [Documenting parameters](http://style.tidyverse.org/code-documentation.html#documenting-parameters):

   For `@param` and `@return`, the text should starting with the expected class (or possible classes) of the argument or return value, followed by a comma, then the (uncapitalized) description. If omitting the class name, then begin the description with a capital letter.

   ```r
   #' @param spec  An object to be coerced to `vegaspec`, a Vega/Vega-Lite specification
   #' @param width `integer`, sets the view width in pixels
   #'
   #' @return `logical` indicating success
   ```

In the documentation, we use *specification* or *spec* to describe the JSON or the list; we use *chart* to describe the rendering, the finished product. These seem to be the terms-of-art that Vega-Lite uses.

## Development strategy

So that we can use the (very useful) pull-request functions from [usethis](https://usethis.r-lib.org/), we follow the Tidyverse convention of using the `master` branch as the reference branch for pull-requests. However, *you* should not make a pull-request from *your* copy of the `master` branch; you should work from a branch named for the change you are proposing. For more information, please see the [usethis pull-request reference](https://usethis.r-lib.org/reference/pr_init.html).

We will wish for `master` to contain only stable versions. We will not normally merge a pull-request that does not pass the CI checks. Further, we will intend that each commit to master will have a incremented version number; we will manage this as a part of the pull-request process.

Please build pkgdown as much as you would like - the `docs` folder is git-ignored; the pkgdown site is built and deployed automatically upon update of the GitHub `master` branch. The CRAN version of the documentation is at the "root" of the documentation site; the latest `master` version will be deployed to the `dev` directory of the "root".

### Pull requests

Pull requests are very welcome. The branch to which you should make a pull-request will depend on the situation:

Situation                  | Reference branch     | Add item to NEWS.md      | Appreciated
-------------------------- | -------------------- | ------------------------ | -----------
bug-fix                    | `master`             | `r mark_yes`             | `r grin`
improving documentation    | `master`             | `r mark_no`              | `r grin`
adding vignette            | `master`             | `r mark_yes`             | `r grin`
helping with a new feature | `<feature-branch>`   | `r mark_no`              | `r grin`
proposing a new feature    | `master`             | `r mark_yes`             | `r grin`

<br>

Please roxygenize as a part of your pull-request. Let's all do our best to keep to the current CRAN verision of roxygen2.

### Versioning

The first digit indicates the maturity of this package's API. For the time being, it will be `0`.

The second digit will be incremented upon each CRAN release and assigned a GitHub release tag.

### Vega versions 

To update the JavaScript files for Vega, Vega-Lite, and vega-embed, a mainainter will render the R markdown document found at `data-raw/infrastructure.Rmd`. The key parameter to adjust is `vega_lite_version`, in the YAML header:

```yaml
---
title: Package infrastrucure
output: github_document
params:
  vega_lite_version: "3.2.1"
---
```

The code in the `.Rmd` file will determine the versions of Vega and vega-embed that are concurrent with this version of Vega-Lite.

When updating the Vega libraries for this package, please keep in mind that you may also have to:

- rebuild & reinstall your new version of this package
- rebuild the datasets (next section)
- rebuild some of the test specs and reference images, found in `tests/spec` and `tests/reference`

One day, this will be better automated.

### Datasets

The sample specifications and datasets are build using the R markdown document found at `data-raw/infrastructure.Rmd`. 

## Development philosophy

### S3 class system

In R, a vegaspec lives as a list, but can be imported from or exported to JSON text. Because a vegaspec is a list, we add S3 classes on top of `list`.

For example:

```{r}
library("vegawidget")
class(spec_mtcars)
```

We see that in addition to list we have classes for:

- `vegaspec`: to include all Vega and Vega-Lite specification
- `vegaspec_vega_lite`: denotes that this is a Vega-Lite specification
- `vegaspec_unit`: denotes that this is a unit specification 

This last (or top) class is used to distinguish [different compositions of Vega-Lite specifications](https://vega.github.io/vega-lite/docs/composition.html). In this case, a unit-specification is one that is a single non-layered view. 

The possible classes are: `vegaspec_unit`, `vegaspec_layer`, `vegaspec_facet`, `vegaspec_repeat`, `vegaspec_concat`, `vegaspec_hconcat`, `vegaspec_vconcat`. 

For Vega specifications, the situation is simpler:

```{r}
class(vw_to_vega(spec_mtcars))
```
