---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-"
)
```

[![lifecycle](https://img.shields.io/badge/lifecycle-maturing-blue.svg)](https://www.tidyverse.org/lifecycle/#maturing)
[![Travis build status](https://travis-ci.org/vegawidget/vegawidget.svg?branch=master)](https://travis-ci.org/vegawidget/vegawidget)
[![CRAN status](https://www.r-pkg.org/badges/version/vegawidget)](https://cran.r-project.org/package=vegawidget)

# vegawidget

The goal of vegawidget is to render Vega-Lite and Vega specifications into htmlwidgets. Its ambition is to be a low-level interface to the Vega(-Lite) API, such that other packages can build upon it to offer higher-level functions to compose Vega(-Lite) specifications. This is the key difference with the [vegalite](https://github.com/hrbrmstr/vegalite) package; this package provides a minimal interface with Vega-Lite. 

## Installation

You can install vegawidget from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("vegawidget/vegawidget")
```

An [interactive tutorial](https://ijlyttle.shinyapps.io/vegawidget-overview) is available at shinyapps.io.
 
## Introduction

Vega(-Lite) specifications are just text, formatted as JSON. However, in R, we can use lists to build such specifications:

```{r spec}
library("vegawidget")

spec_mtcars <-
  list(
    `$schema` = vega_schema(), # specifies Vega-Lite
    description = "An mtcars example.",
    data = list(values = mtcars),
    mark = "point",
    encoding = list(
      x = list(field = "wt", type = "quantitative"),
      y = list(field = "mpg", type = "quantitative"),
      color = list(field = "cyl", type = "nominal")
    )
  ) %>% 
  as_vegaspec()
```

The `as_vegaspec()` function is used to turn the list into a *vegaspec*; a good many of the functions in this package are built to support, and render, vegaspecs: 

```{r vegawidget}
spec_mtcars
```

If you are reading this document on the GitHub code site, it is further rendered into a PNG. In the background, the vegaspec print-method converts the vegaspec to JSON, using `vw_as_json()`, then renders it as an htmlwidget using `vegawidget()`.

If you wish to examine a vegaspec, you may use the `vw_examine()` function, which is a thin wrapper around the `jsonedit()` function from the **listviewer** package.

```{r}
vw_examine(spec_mtcars)
```

If you wish to deploy a vegaspec to be rendered on [bl.ocks.org](https://bl.ocks.org/), you can do that, provided you have a GitHub account.

```{r create_block, eval=FALSE}
vw_create_block(spec_mtcars)
```

Similarly, if a block has a `.json` file with a JSON specification, you can retrieve and put into a vegaspec:

```{r retrieve_block, eval=FALSE}
spec_new <- vw_retrieve_block("ijlyttle/06a512525fbe7a96e1a2028e1871b61c")
```

## Integration with other packages

This package provides functions to render Vega(-Lite) specifications; although it provides some helpers, it dos not provide higher-level functions to build specifications. Rather, this is left to other packages. Even though you can use its functions directly, you are invited to import and re-export them for use in *your* package. 

Accordingly, this package offers a templating function, `use_vegawidget()`, to help you integrate vegawidget functions into your package. This function is used to [import and re-export vegawidget functions](https://github.com/vegawidget/altair/blob/master/R/utils-vegawidget.R) for the [altair](https://vegawidget.github.io/altair) package.
 
## Acknowlegdements

- [Alicia Schep](https://github.com/AliciaSchep) has been instrumental in guiding the evolution of the API.
- [Bob Rudis](https://github.com/hrbrmstr) and the [vegalite](https://github.com/hrbrmstr/vegalite) package provided a lot of the inspiration for this work, providing a high-level interface to Vega-Lite.
- The [Altair](https://altair-viz.github.io) developers, for further popularizing the notion of using a programming language (Python) to create and render Vega-Lite specifications.  
- The [Vega-Lite](https://vega.github.io/vega-lite/) developers, for providing a foundation upon which the rest of this is built. 

## Contributing

Contributions are welcome, please see this [guide](CONTRIBUTING.md).
Please note that this project is released with a [Contributor Code of Conduct](CODE_OF_CONDUCT.md). By participating in this project you agree to abide by its terms.


