---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![lifecycle](https://img.shields.io/badge/lifecycle-experimental-orange.svg)](https://www.tidyverse.org/lifecycle/#experimental)
[![Travis build status](https://travis-ci.org/vegawidget/vegawidget.svg?branch=master)](https://travis-ci.org/vegawidget/vegawidget)
[![CRAN status](https://www.r-pkg.org/badges/version/vegawidget)](https://cran.r-project.org/package=vegawidget)

# vegawidget

The goal of vegawidget is to render Vega-Lite and Vega specifications into htmlwidgets. For now, this document serves as a manifesto for what we might hope for the package.

```{r vegawidget}
library("vegawidget")

spec_mtcars <-
  list(
    `$schema` = "https://vega.github.io/schema/vega-lite/v2.json",
    description = "An mtcars example.",
    data = list(values = mtcars),
    mark = "point",
    encoding = list(
      x = list(field = "wt", type = "quantitative"),
      y = list(field = "mpg", type = "quantitative"),
      color = list(field = "cyl", type = "nominal")
    )
  )  

vegawidget(spec_mtcars)
```

```{r}
examine(spec_mtcars)
```

## Specifications

Vega and Vega-Lite specifications are just text, formatted as JSON. It is this package's responsibiity to render such specifications; allowing you to develop packages that *build* specifications. To take advantage of this package, it will be easiest if your specification built by your package is an S3 class based on lists or JSON.

For example, consider the **altair** package, which uses **reticulate** to work with the **Altair** Python package to build Vega-Lite specifications.

```r
class(altair::alt$Chart())
```

```
[1] "altair.vegalite.v2.api.Chart"                          
[2] "altair.vegalite.v2.api.TopLevelMixin"                  
[3] "altair.vegalite.v2.schema.mixins.ConfigMethodMixin"    
[4] "altair.vegalite.v2.api.EncodingMixin"                  
[5] "altair.vegalite.v2.schema.mixins.MarkMethodMixin"      
[6] "altair.vegalite.v2.schema.core.TopLevelFacetedUnitSpec"
[7] "altair.vegalite.v2.schema.core.VegaLiteSchema"         
[8] "altair.utils.schemapi.SchemaBase"                      
[9] "python.builtin.object" 
```

The 

The central function here will be `vegawidget()`, to render an htmlwidget from a Vega/Vega-Lite specification. Supporting functions will include:

- `examine()` 
- `create_block()`

Behind the scenes, the workhorse function is `as_vegaspec()`. To implement the `vegawidget()` functions for "your" S3 object that implements a Vega or Vega-Lite specification, you may have to implement a method for your class for the `as_vegaspec()` generic, as well as methods for `print()` and `knit_print()`.

The vegawidget 

- `list`
- `character` 
- **jsonlite** `json` object 

Note about webshot and phantomJS.

For a package such as **altair**, it becomes that pacakge's responsibility to provide methods for its objects.

In the future, challenges will include:

- when knitting to a non-interactive format, e.g. `github_document` getting the "right" thing to happen
- offering **crosstalk** compatibility
- offering reactive **shiny** behavior, like **ggvis** provides for its Vega specifications

## Installation

You can install vegawidget from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("vegawidget/vegawidget")
```


