---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

[![Travis build status](https://travis-ci.org/ijlyttle/vegawidget.svg?branch=master)](https://travis-ci.org/ijlyttle/vegawidget)

# vegawidget

The goal of vegawidget is to handle the rendering into htmlwidgets for Vega-Lite and Vega specifications.

For now, this document will serve a manifesto for what we might hope for the package.

## Specifications

The beautiful thing about Vega and Vega-Lite specifications is that they are just text, formatted as JSON.

The central function here will be `vegawidget()`. Its supporting function will be `examine()`.

I think each of these functions should be S3 generics for which we provide methods for:

- list
- character 
- `jsonlite` `json` object 

For a package such as **altair**, it becomes that packge's responsibility to provide generics for its objects.

In the future, challenges will include:

- using `gistr` to create a block.
- when knitting to a non-interactive format, e.g. `github_document` getting the "right" thing to happen
- offering **crosstalk** compatibility
- offering reactive **shiny** behavior, like **ggvis** provides for its Vega specifications

## Installation

You can install vegawidget from github with:

```{r gh-installation, eval = FALSE}
# install.packages("devtools")
devtools::install_github("vegawidget/vegawidget")
```


