---
title: htmlwidgets lib files
output: github_document
params:
  vega_lite_version: "2.5.0"
---

The purpose of this document is to assemble the files for the `inst/htmlwidgets` directory.

To construct the `inst/htmlwidgets` directory, we have two "sources of truth": this file and the contents of the directory `data-raw/templates/vegawidget`.

The **only** way to put anything in the `inst/htmlwidgets` directory is to run this file.

## Configure

```{r packages}
library("conflicted")
library("fs")
library("glue")
library("httr")
library("here")
library("tibble")
library("purrr")
library("readr")
library("vegawidget")
```

```{r}
dir_htmlwidgets <- here("inst", "htmlwidgets")
dir_templates <- here("data-raw", "templates")

vega_versions_long <- vega_versions(params$vega_lite_version)

# we want to remove the "-rc.2" from the end of "4.0.0-rc.2"
# "-\\w.*$"   hyphen, followed by a letter, followed by anything, then end 
vega_versions_short <- map(vega_versions_long, ~sub("-\\w.*$", "", .x))
```

## Clean and create

Our first task is to create a clean directory `inst/htmlwidgets`.

```{r}
params$dir_htmlwidgets
```

```{r clean-create}
if (dir_exists(params$dir_htmlwidgets)) {
  dir_delete(params$dir_htmlwidgets)
}

dir_create(params$dir_htmlwidgets)
```

## Vegawidget files

We have a couple of files to copy from our templates directory.

```{r vegawidget-js}
file_copy(
  path(dir_templates, "vegawidget.js"), 
  path(dir_htmlwidgets, "vegawidget.js")
)
```

```{r vegawidget-yml}
path(dir_templates, "vegawidget.yaml") %>%
  read_lines() %>%
  map_chr(~glue_data(vega_versions_short, .x)) %>%
  write_lines(path(dir_htmlwidgets, "vegawidget.yaml"))
```

## Lib directory
