---
title: htmlwidgets lib files
output: github_document
params:
  vega_lite_version: "2.5.0"
---

This document is run **only** if you want to reassemble the vegawidget htmlwidget. This would be done to update this package in whenever the **Vega-Lite** version is upgraded.

In concrete terms, this document will construct the `inst/htmlwidgets` directory; the **only** way to put anything in the `inst/htmlwidgets` directory is to run this file.

We have two "sources of truth": this file and the contents of the directory `data-raw/templates`.

As long as the vega-embed framework remains the same, all we should need to do to update the **Vega-Lite** libraries is to change the parameter `vega_lite_version` in the yaml-header to this file.

If you want to update the workings of the htmlwidget, you will have to do so in the files `data-raw/templates` directory. It may be useful to note this in a "contributing" document for this package.

## Configure

These packages are not listed in the `Suggests` section of the `DESCRIPTION` file. It's on you to make sure they are all up-to-date.

```{r packages}
library("conflicted")
library("fs")
library("glue")
library("httr")
library("here")
library("tibble")
library("purrr")
library("readr")
library("dplyr")
library("vegawidget")
```

For this document, there is a source-directory, `dir_templates`, and a target-directory, `dir_htmlwidgets`. We will read information from the source-directory and write it to the target-directory. 

```{r directories}
dir_templates <- here("data-raw", "templates")
dir_htmlwidgets <- here("inst", "htmlwidgets")
```

Finally, we need to know which versions of the libraries (vega, vega-lite, and vega-embed) to download. We do this by inspecting the manifest of a specific version of the vega-lite library. This package has an internal function, `vega_version()` to help us do this:  

```{r versions}
vega_versions_long <- vega_versions(params$vega_lite_version)

vega_versions_long

# we want to remove the "-rc.2" from the end of "4.0.0-rc.2"
# "-\\w.*$"   hyphen, followed by a letter, followed by anything, then end 
vega_versions_short <- map(vega_versions_long, ~sub("-\\w.*$", "", .x))
```

## Clean and create

Our first task is to create a clean directory `inst/htmlwidgets`. If it exists, we delete it and create it anew.

```{r clean-create}
if (dir_exists(dir_htmlwidgets)) {
  dir_delete(dir_htmlwidgets)
}

dir_create(dir_htmlwidgets)
```

## Vegawidget files

Here, we have copy some files from our templates directory.

```{r vegawidget-js}
file_copy(
  path(dir_templates, "vegawidget.js"), 
  path(dir_htmlwidgets, "vegawidget.js")
)
```

The file `vegawidget.yaml` requires the versions the JavaScript libraries; we interpolate these from `vega_versions_short`.

```{r vegawidget-yml}
path(dir_templates, "vegawidget.yaml") %>%
  read_lines() %>%
  map_chr(~glue_data(vega_versions_short, .x)) %>%
  write_lines(path(dir_htmlwidgets, "vegawidget.yaml"))
```

## Lib directory

Here's where we download the libraries themselves, along with the licences; the versions are interpolated from `vega_versions_long`.

```{r}
downloads <-
  tribble(
    ~path_local,                         ~path_remote,
    "vega-lite/vega-lite-min.js",        "https://cdn.jsdelivr.net/npm/vega-lite@{vega_lite}",
    "vega-lite/LICENSE",                 "https://raw.githubusercontent.com/vega/vega-lite/master/LICENSE",
    "vega/promise.min.js",               "https://vega.github.io/vega/assets/promise.min.js",
    "vega/symbol.min.js",                "https://vega.github.io/vega/assets/symbol.min.js",
    "vega/vega.js",                      "https://cdn.jsdelivr.net/npm/vega@{vega}/build/vega.js",
    "vega/LICENSE",                      "https://raw.githubusercontent.com/vega/vega/master/LICENSE",
    "vega-embed/vega-embed.js",          "https://cdn.jsdelivr.net/npm/vega-embed@{vega_embed}",
    "vega-embed/LICENSE",                "https://raw.githubusercontent.com/vega/vega-embed/master/LICENSE"
  ) %>%
  mutate(
    path_remote = map_chr(path_remote, ~glue_data(vega_versions_long, .x))
  ) 
```

```{r}
get_file <- function(path_local, path_remote, lib_dir) {
  
  path_local <- fs::path(lib_dir, path_local)
  
  # if directory does not yet exist, create it
  dir_local <- fs::path_dir(path_local)
  
  if (!fs::dir_exists(dir_local)) {
    dir_create(dir_local)
  }
  
  resp <- httr::GET(path_remote)
  
  text <- httr::content(resp, type = "text", encoding = "UTF-8")
  
  readr::write_file(text, path_local)
  
  invisible(NULL)
}
```

Here, we create the `lib` directory, then "walk" through each row of the `downloads` data frame to get each of the files and put it into place.

```{r}
dir_lib <- path(dir_htmlwidgets, "lib")
dir_create(dir_lib)

pwalk(downloads, get_file, lib_dir = dir_lib)
```

## Patch 

Here, Alicia Schep noticed that there was a problem to render vega charts within the RStudio IDE, and she figured out a workaround (as well as a [PR]() for the RStudio IDE to fix the problem). Here's her patch for older versions of the IDE: 

```{r}
vega_embed_path <- path(dir_lib, "vega-embed/vega-embed.js")
vega_embed <- readr::read_file(vega_embed_path)

vega_mod <- stringr::str_replace_all(vega_embed, 'head>"','he"+"ad>"') 
vega_mod <- stringr::str_replace_all(vega_mod, '"<\\/head>','"</he"+"ad>') 

readr::write_file(vega_mod, path(dir_lib, "vega-embed/vega-embed-modified.js"))
fs::file_delete(vega_embed_path)
```
