---
title: Package infrastrucure
output: github_document
params:
  versions: 
    value:
      vl5: "5.5.0"
      vl4: "4.17.0"
  widget_default: "vl5"
---

This purpose of this document is to build the package infrastucture. 

To upgrade the versions of **Vega-Lite** that we support:

1. Modify the parameter `versions$value` in the yaml-header to this file.
2. Render (knit) this document.
3. Build-and-install this package on your local computer.
4. Run the tests (`testthat::test()`); some snapshots will have changed.
5. Rebuild the pkgdown website (`pkgdown::build_site()`), verify the visual-regression article (still to be built).
6. Commit, push, and make PR.

These packages may not be listed in the `Suggests` section of the `DESCRIPTION` file. It's on you to make sure they are all up-to-date:

```{r packages}
library("fs")
library("glue")
library("here")
library("purrr")
library("readr")
library("dplyr")
library("tibble")
library("stringr")
library("usethis")
library("conflicted")
library("vegawidget")

conflict_prefer("filter", "dplyr")
```

## Infrastructure

There are two source of "truth" for this process: this document, and the contents of the directory `data-raw/templates`. 
(It may be useful to note this in a "contributing" document for this package.)
Each of the infrastructure elements is created anew when this document is run; so, between this document and `data-raw/templates`, we need to be able to construct completely all the infrastructure.

Thus, we define this template directory and a function to delete and create 

```{r directories}
dir_templates <- here("data-raw", "templates")

# create a create a clean directory, with a safety
create_clean <- function(path, path_safe = here::here()) {
  
  # tests that path is within path_safe
  is_path_safe <- function(path, path_safe) {
    identical(
      unclass(path_safe), 
      unclass(fs::path_common(c(path, path_safe)))
    )
  }
  
  if (!is_path_safe(path, path_safe)) {
    stop(
      "Nothing done because ",
      " path: ", shQuote(path), 
      " is not within path_safe: ", shQuote(path_safe)
    )
  }
  
  if (fs::dir_exists(path)) {
    fs::dir_delete(path)
  }
  
  fs::dir_create(path)
  
}
```

## Configure

We need to know which versions of the libraries (vega, vega-lite, and vega-embed) to download. We do this by inspecting the manifest of a specific version of the vega-lite library. This package has an internal function, `get_vega_version()` to help us do this:  

```{r}
tbl_versions <- 
  enframe(unlist(params$versions), name = "widget", value = "vega_lite")
```

```{r version-multi}
vega_version_all <- 
  tbl_versions |>
  left_join(
    map_dfr(tbl_versions$vega_lite, vegawidget:::get_vega_version),
    by = "vega_lite"
  )
   
vega_version_all

# we want to remove the "-rc.2" from the end of "4.0.0-rc.2"
# "-\\w.*$"   hyphen, followed by a letter, followed by anything, then end 
vega_version_short_all <- 
  vega_version_all |>
  mutate(across(starts_with("vega"), ~sub("-\\w.*$", "", .x)))
  
vega_version_short_all

if (!identical(vega_version_all, vega_version_short_all)) {
  warning("You may be using a release-candidate; please check.")
}
```

## htmlwidgets 

First, let's create a clean directory for the htmlwidget

```{r vegawidget-dir}
dir_htmlwidgets <- here("inst", "htmlwidgets")
dir_lib <- path(dir_htmlwidgets, "lib")
dir_vegaembed <- path(dir_lib, "vega-embed")

create_clean(dir_htmlwidgets)
dir_create(dir_lib)
dir_create(dir_vegaembed)
```

### vegawidget

First, copy some files from our templates directory.

```{r vegawidget-js-multi}
write_js <- function(x) {
  path(dir_templates, "vegawidget-all.js") %>%
    readr::read_lines() %>%
    purrr::map_chr(
      ~glue::glue_data(list(widget = x), .x, .open = "${")
    ) %>%
    readr::write_lines(path(dir_htmlwidgets, glue::glue("vegawidget-{x}.js")))  
}

walk(tbl_versions$widget, write_js)
```

The file `vegawidget-all.yaml` requires the versions the JavaScript libraries; we interpolate these from `vega_version_short`.

```{r vegawidget-yml-multi}
write_yaml <- function(x) {
  path(dir_templates, "vegawidget-all.yaml") %>%
    readr::read_lines() %>%
    purrr::map_chr(
      ~glue::glue_data(vega_version_all %>% filter(widget == !!x), .x)
    ) %>%
    readr::write_lines(path(dir_htmlwidgets, glue::glue("vegawidget-{x}.yaml")))  
}

walk(vega_version_all$widget, write_yaml)
```

The file `vega-embed.css` adds some css for the (old-style) links that appeared below a rendered spec:

```{r}
fs::file_copy(
  fs::path(dir_templates, "vega-embed.css"), 
  fs::path(dir_vegaembed, "vega-embed.css")
)
```

Here's where we download the libraries themselves, along with the licences.

```{r}
license_downloads <- 
    tribble(
    ~path_local,                         ~path_remote,
    "vega-lite/LICENSE",                 "https://raw.githubusercontent.com/vega/vega-lite/next/LICENSE",
    "vega/LICENSE",                      "https://raw.githubusercontent.com/vega/vega/master/LICENSE",
    "vega-embed/LICENSE",                "https://raw.githubusercontent.com/vega/vega-embed/next/LICENSE"
  )

license_downloads
```

```{r}
downloads_template <- 
  tribble(
    ~path_local,                                 ~path_remote,
    "vega-lite/vega-lite@{vega_lite}.min.js",    "https://cdn.jsdelivr.net/npm/vega-lite@{vega_lite}",
    "vega/vega@{vega}.min.js",                   "https://cdn.jsdelivr.net/npm/vega@{vega}",
    "vega-embed/vega-embed@{vega_embed}.min.js", "https://cdn.jsdelivr.net/npm/vega-embed@{vega_embed}",
  )

downloads_local <- function(widget) {
  
  df_local <- dplyr::filter(vega_version_all, widget == !!widget)
  
  downloads_template |>
  dplyr::mutate(
    dplyr::across(
      dplyr::starts_with("path"), 
      ~map_chr(.x, ~glue::glue_data(df_local, .x))
    )
  )  
}

htmlwidgets_downloads_all <- map_dfr(vega_version_all$widget, downloads_local)

htmlwidgets_downloads_all
```

```{r}
get_file <- function(path_local, path_remote, path_local_root) {
  
  path_local <- fs::path(path_local_root, path_local)
  
  # if directory does not yet exist, create it
  # TODO: swap out with create_clean()
  dir_local <- fs::path_dir(path_local)
  
  if (!fs::dir_exists(dir_local)) {
    dir_create(dir_local)
  }
  
  download.file(path_remote, destfile = path_local)

  invisible(NULL)
}
```

Here, we create the `lib` directory, then "walk" through each row of the `downloads` data frame to get each of the files and put it into place.

```{r}
pwalk(license_downloads, get_file, path_local_root = dir_lib)
```

```{r}
pwalk(htmlwidgets_downloads_all, get_file, path_local_root = dir_lib)
```

### Versions

We use this to support the `vega_version()` function.

```{r}
.vega_version_all <- as.data.frame(vega_version_all)
.widget_default <- params$widget_default
```

### Vegawidget handlers

```{r}
.vw_handler_library <-
  list(
    event = .vw_handler_def(
      args = c("event", "item"),
      bodies = list(
        item = .vw_handler_body(
          params = list(),
          text = c(
            "// returns the item",
            "return item;"
          )
        ),
        datum = .vw_handler_body(
          params = list(),
          text = c(
            "// if there is no item or no datum, return null",
            "if (item === null || item === undefined || item.datum === undefined) {",
            "  return null;",
            "}",
            "",
            "// returns the datum behind the mark associated with the event",
            "return item.datum;"
          )
        )
      )
    ),
    signal = .vw_handler_def(
      args = c("name", "value"),
      bodies = list(
        value = .vw_handler_body(
          params = list(),
          text = c(
            "// returns the value of the signal",
            "return value;"
          )
        )
      )
    ),
    data = .vw_handler_def(
      args = c("name", "value"),
      bodies = list(
        value = .vw_handler_body(
          params = list(),
          text = c(
            "// returns a copy of the data",
            "return value.slice();"
          )
        )
      )
    ),
    effect = .vw_handler_def(
      args = "x",
      bodies = list(
        shiny_input = .vw_handler_body(
          params = list(inputId = NULL, expr = "x"),
          text = c(
            "// sets the Shiny-input named `inputId` to `expr` (default \"x\")",
            "Shiny.setInputValue('${inputId}', ${expr});"
          )
        ),
        console = .vw_handler_body(
          params = list(label = "", expr = "x"),
          text = c(
            "// if `label` is non-trivial, prints it to the JS console",
            "'${label}'.length > 0 ? console.log('${label}') : null;",
            "// prints `expr` (default \"x\") to the JS console",
            "console.log(${expr});"
          )
        ),        
        element_text = .vw_handler_body(
          params = list(selector = NULL, expr = "x"),
          text = c(
            "// to element `selector`, adds text `expr` (default \"x\")",
            "document.querySelector('${selector}').innerText = ${expr}"
          )
        )
      )
    )
  )
```

### Write this out

```{r}
usethis::use_data(
  .vega_version_all,
  .widget_default,
  .vw_handler_library,
  internal = TRUE, 
  overwrite = TRUE
)
```

