---
title: Package infrastrucure
output: github_document
params:
  vega_lite_version: "4.0.2"
---

This purpose of this document is to build the package infrastucture. 

To upgrade the version of **Vega-Lite** that we support:

1. Modify the parameter `vega_lite_version` in the yaml-header to this file.
2. Render (knit) this document.
3. Build-and-install this package on your local computer.
4. Run the tests (`testthat::test()`).
5. Rebuild the pkgdown website (`pkgdown::build_site()`), verify the visual-regression article (still to be built).
6. Commit, push, and make PR.

```{r packages}
library("fs")
library("glue")
library("httr")
library("here")
library("purrr")
library("readr")
library("dplyr")
library("tibble")
library("stringr")
library("usethis")
library("conflicted")
library("vegawidget")
```

## Infrastructure

Package infrastucture incudes:

- an htmlwidget named "vegawidget"
- internal package data: 
  - list of version numbers: `.vega_version`
- files to validate the schema

Perhaps this could be a series of documents - it remains as an exercise to see what can be cleaved away.

There are two source of "truth" for this process: this document, and the contents of the directory `data-raw/templates`. (It may be useful to note this in a "contributing" document for this package.)
Each of the infrastructure elements is created anew when this document is run; so, between this document and `data-raw/templates`, we need to be able to construct completely each element infrastructure.

Thus, we define this template directory and a function to delete and create 

```{r directories}
dir_templates <- here("data-raw", "templates")

# create a create a clean directory, with a safety
create_clean <- function(path, path_safe = here::here()) {
  
  # tests that path is within path_safe
  is_path_safe <- function(path, path_safe) {
    identical(
      unclass(path_safe), 
      unclass(fs::path_common(c(path, path_safe)))
    )
  }
  
  if (!is_path_safe(path, path_safe)) {
    stop(
      "Nothing done because ",
      " path: ", shQuote(path), 
      " is not within path_safe: ", shQuote(path_safe)
    )
  }
  
  if (fs::dir_exists(path)) {
    fs::dir_delete(path)
  }
  
  fs::dir_create(path)
  
}
```

## Configure

These packages are not listed in the `Suggests` section of the `DESCRIPTION` file. It's on you to make sure they are all up-to-date.

We need to know which versions of the libraries (vega, vega-lite, and vega-embed) to download. We do this by inspecting the manifest of a specific version of the vega-lite library. This package has an internal function, `vega_version()` to help us do this:  

```{r version}
vega_version_long <- vegawidget:::get_vega_version(params$vega_lite_version)

vega_version_long

# we want to remove the "-rc.2" from the end of "4.0.0-rc.2"
# "-\\w.*$"   hyphen, followed by a letter, followed by anything, then end 
vega_version_short <- map(vega_version_long, ~sub("-\\w.*$", "", .x))
```

## htmlwidgets 

First, let's create a clean directory for the htmlwidget

```{r vegawidget-dir}
dir_htmlwidgets <- here("inst", "htmlwidgets")
dir_lib <- path(dir_htmlwidgets, "lib")
dir_vegaembed <- path(dir_lib, "vega-embed")

create_clean(dir_htmlwidgets)
dir_create(dir_lib)
dir_create(dir_vegaembed)
```

### vegawidget

First, copy some files from our templates directory.

```{r vegawidget-js}
file_copy(
  path(dir_templates, "vegawidget.js"), 
  path(dir_htmlwidgets, "vegawidget.js"),
  overwrite = TRUE # take this out when done
)
```

The file `vegawidget.yaml` requires the versions the JavaScript libraries; we interpolate these from `vega_version_short`.

```{r vegawidget-yml}
path(dir_templates, "vegawidget.yaml") %>%
  read_lines() %>%
  map_chr(~glue_data(vega_version_short, .x)) %>%
  write_lines(path(dir_htmlwidgets, "vegawidget.yaml"))
```

The file `vega-embed.css` adds some css for the (old-style) links that appeared below a rendered spec:

```{r}
fs::file_copy(
  fs::path(dir_templates, "vega-embed.css"), 
  fs::path(dir_vegaembed, "vega-embed.css")
)
```

Here's where we download the libraries themselves, along with the licences; the versions are interpolated from `vega_version_long`.

```{r}
htmlwidgets_downloads <-
  tribble(
    ~path_local,                         ~path_remote,
    "vega-lite/vega-lite.min.js",        "https://cdn.jsdelivr.net/npm/vega-lite@{vega_lite}",
    "vega-lite/LICENSE",                 "https://raw.githubusercontent.com/vega/vega-lite/master/LICENSE",
    "vega/vega.min.js",                  "https://cdn.jsdelivr.net/npm/vega@{vega}",
    # "vega/vega.js",                      "https://cdn.jsdelivr.net/npm/vega@{vega}/build/vega.js",
    "vega/LICENSE",                      "https://raw.githubusercontent.com/vega/vega/master/LICENSE",
    "vega-embed/vega-embed.js",          "https://cdn.jsdelivr.net/npm/vega-embed@{vega_embed}",
    "vega-embed/LICENSE",                "https://raw.githubusercontent.com/vega/vega-embed/master/LICENSE"
  ) %>%
  mutate(
    path_remote = map_chr(path_remote, ~glue_data(vega_version_long, .x))
  )

htmlwidgets_downloads
```

```{r}
get_file <- function(path_local, path_remote, path_local_root) {
  
  path_local <- fs::path(path_local_root, path_local)
  
  # if directory does not yet exist, create it
  # TODO: swap out with create_clean()
  dir_local <- fs::path_dir(path_local)
  
  if (!fs::dir_exists(dir_local)) {
    dir_create(dir_local)
  }
  
  resp <- httr::GET(path_remote)
  
  text <- httr::content(resp, type = "text", encoding = "UTF-8")
  
  readr::write_file(text, path_local)
  
  invisible(NULL)
}
```

Here, we create the `lib` directory, then "walk" through each row of the `downloads` data frame to get each of the files and put it into place.

```{r}
pwalk(htmlwidgets_downloads, get_file, path_local_root = dir_lib)
```

### Patch 

Here, Alicia Schep noticed that there was a problem to render vega charts within the RStudio IDE, and she figured out a workaround (as well as a PR for the RStudio IDE to fix the problem). Here's her patch for older versions of the IDE: 

```{r eval=FALSE}
# disabling for now - to see if this works without the patch
vega_embed_path <- path(dir_lib, "vega-embed/vega-embed.js")
vega_embed <- readr::read_file(vega_embed_path)

vega_mod <- stringr::str_replace_all(vega_embed, 'head>"','he"+"ad>"') 
vega_mod <- stringr::str_replace_all(vega_mod, '"<\\/head>','"</he"+"ad>') 

readr::write_file(vega_mod, path(dir_lib, "vega-embed/vega-embed-modified.js"))
fs::file_delete(vega_embed_path)
```

## Schema

One of the purposes of this package is to provide a means to validate a spec. 

```{r vegawidget-schema}
dir_schema <- here("inst", "schema")
create_clean(dir_schema)
```

Having thought about this (perhaps too much), the only reasonable way to go for a given release to support only a single version of the javascript libraries and the schema.

```{r}
schema <- 
  tribble(
    ~path_local,                   ~path_remote,
    "vega/v{vega}.json",           "https://vega.github.io/schema/vega/v{vega}.json",
    "vega-lite/v{vega_lite}.json", "https://vega.github.io/schema/vega-lite/v{vega_lite}.json"
  ) %>%
  mutate(
    path_local = map_chr(path_local, ~glue_data(vega_version_long, .x)),
    path_remote = map_chr(path_remote, ~glue_data(vega_version_long, .x))
  )

schema
```

```{r}
pwalk(schema, get_file, path_local_root = dir_schema)
```

We want to add a newline to the end of each of these files.

```{r}
walk(
  schema$path_local,
  ~write("\n", file = file.path(dir_schema, .x), append = TRUE)
)
```

### Versions

We use this to support the `vega_version()` function.

```{r}
.vega_version <- vega_version_long
```

### JavaScript Libraries

We want to keep copies of the Vega and Vega-Lite specs so that we can use them with V8.

Another way to do this may be to keep memoised functions that read the files kept in the htmlwidgets directory. This seems potentially cleaner to execute, but more-complicated to implement.

Let look at `htmlwidgets_downloads`

```{r}
regex <- "vega(-lite)?/(.*)\\.min.js$"

htmlwidgets_vegajs <-
  htmlwidgets_downloads %>%
  dplyr::filter(str_detect(path_local, regex)) %>%
  dplyr::mutate(
    name = str_replace(path_local, regex, ".\\2_js") %>% str_replace("-", "_")
  ) %>%
  dplyr::select(name, path_local)

htmlwidgets_vegajs  
```

We need to put these into the local environment. This smells like a side-effect.

```{r}
assign_js <- function(name, path_local, path_root) {
  
  js <- readr::read_file(fs::path(path_root, path_local))
  
  assign(name, js, envir = globalenv())
  NULL 
}

pwalk(htmlwidgets_vegajs, assign_js, path_root = dir_lib)
```

Vegawidget handlers:

```{r}
.vw_handler_library <-
  list(
    event = .vw_handler_def(
      args = c("event", "item"),
      bodies = list(
        item = .vw_handler_body(
          params = list(),
          text = c(
            "// returns the item",
            "return item;"
          )
        ),
        datum = .vw_handler_body(
          params = list(),
          text = c(
            "// if there is no item or no datum, return null",
            "if (item === null || item === undefined || item.datum === undefined) {",
            "  return null;",
            "}",
            "",
            "// returns the datum behind the mark associated with the event",
            "return item.datum;"
          )
        )
      )
    ),
    signal = .vw_handler_def(
      args = c("name", "value"),
      bodies = list(
        value = .vw_handler_body(
          params = list(),
          text = c(
            "// returns the value of the signal",
            "return value;"
          )
        )
      )
    ),
    data = .vw_handler_def(
      args = c("name", "value"),
      bodies = list(
        value = .vw_handler_body(
          params = list(),
          text = c(
            "// returns a copy of the data",
            "return value.slice();"
          )
        )
      )
    ),
    effect = .vw_handler_def(
      args = "x",
      bodies = list(
        shiny_input = .vw_handler_body(
          params = list(inputId = NULL, expr = "x"),
          text = c(
            "// sets the Shiny-input named `inputId` to `expr` (default \"x\")",
            "Shiny.setInputValue('${inputId}', ${expr});"
          )
        ),
        console = .vw_handler_body(
          params = list(label = "", expr = "x"),
          text = c(
            "// if `label` is non-trivial, prints it to the JS console",
            "'${label}'.length > 0 ? console.log('${label}') : null;",
            "// prints `expr` (default \"x\") to the JS console",
            "console.log(${expr});"
          )
        ),        
        element_text = .vw_handler_body(
          params = list(selector = NULL, expr = "x"),
          text = c(
            "// to element `selector`, adds text `expr` (default \"x\")",
            "document.querySelector('${selector}').innerText = ${expr}"
          )
        )
      )
    )
  )
```


```{r}
usethis::use_data(
  .vega_version, 
  .vw_handler_library,
  internal = TRUE, 
  overwrite = TRUE
)
```

