---
title: "Share with bl.ocks.org"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

```{r}
library("vegawidget")
```

Mike Bostock's [blocks service](https://bl.ocks.org) is a great resource for sharing visualizations. *Blocks* (as they are called) are not persistenly stored on the blocks website; they are taken from GitHub [gists](https://gist.github.com).

- If you have the [**gistr**](https://cran.r-project.org/package=gistr) and [**fs**](https://fs.r-lib.org) packages installed, and have a GitHub account and personal-access token (PAT), you can deploy a `vegaspec` to gist, and it will appear as a block.

- If you have the [**magick**](https://github.com/ropensci/magick#readme) and [**webdriver**](https://github.com/rstudio/webdriver#README) packages (with PhantomJS installed), you can include preview and thumb images with the block.

- If a block (not necessarily yours) has a Vega(-Lite) specification, and the specification is in its own file, you can import a `vegaspec` from a block. 

As you may know, it can be a frustrating experience to work with git from R - until it isn't. By this, I mean that you can spend a lot of energy to get things working the first time - then they "just work". If this is your situation, and you have not yet reached the "just work" stage, I highly recommend Jenny Bryan's e-book [Happy Git and GitHub for the useR](http://happygitwithr.com/) to help you get to the essence of git and GitHub issues. If it's a problem, chances are that Jenny has seen it and documented it!

In particular, I have run into issues with SSH credentialing - for which this [troubleshooting section](http://happygitwithr.com/ssh-keys.html#ssh-troubleshooting) is invaluable. 

To use these functions to create and retrieve blocks, you will need a GitHub PAT. Unsurprisingly, there's a book [section](http://happygitwithr.com/github-pat.html) on GitHub PATs.

## Create a block

You can create a block using a vegaspec as an argument to `vw_create_block()`.

```{r eval=FALSE}
vw_create_block(spec_mtcars)
```

```{r echo=FALSE}
message(
  "block url: https://bl.ocks.org/06a512525fbe7a96e1a2028e1871b61c\n",
  " gist url: https://gist.github.com/06a512525fbe7a96e1a2028e1871b61c"
)
```

Here's a [link](https://bl.ocks.org/06a512525fbe7a96e1a2028e1871b61c) to the deployed block.

### Arguments

As you can see in the reference for the `vw_create_block()` function, there are a lot of arguments. Here are some of the highlights:

- `embed` to supply vega-embed options, using the helper-function `vega_embed()`.

- `.block` to supply block-configuration options, using the helper-function `vw_block_config()`. 

- `version` is used to determine if the block uses the versions vega libraries associated with this release of this (vegawidget) package, or if they should use the major-release version. The default is to use the same version.

- `description` is a text string used as a title. If none is supplied here, the function will try to look for a `description` element in the supplied vegaspec.

- `readme` text for a Markdown file, or the path to a Markdown file.

### Return value

We have two very similar functions `vw_create_block()` and `vw_create_block_gistid()`. They have the same arguments and they have the same side-effect; they both create the same gist that can be used as a block. They  differ in the return value:

- `vw_create_block()` returns an invisible copy of the `spec` argument.

- `vw_create_block_gistid()` returns a list with the `endpoint` of the GitHub service used, and the `id` of the gist created.

Depending on what you are doing, one function may be more-useful to you than the other.

```{r eval=FALSE}
gistid <- vw_create_block_gistid(spec_mtcars)
```
```{r echo=FALSE}
message(
  "block url: https://bl.ocks.org/650e59f6a81cf63b9004ae291fe90667\n",
  " gist url: https://gist.github.com/650e59f6a81cf63b9004ae291fe90667"
)

gistid <- list(
  endpoint = "https://api.github.com",
  id = "650e59f6a81cf63b9004ae291fe90667"
)
```

```{r}
gistid
```

## Retrieve a block

If a block has a JSON file, you can retrieve that JSON file and parse it into a vegaspec. For example, we can retrieve the vegaspec from the block that created above:

```{r}
vw_retrieve_block(gistid$id)
```

The default behavior is to return the first file in the gist that has a `.json` extension. If you need to be more specific, you can use the `file` argument.


