---
title: "Share with bl.ocks.org"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

Mike Bostock's [blocks service](https://bl.ocks.org) is a great resource for sharing visualizations. *Blocks* (as they are called) are not stored at this website, they are taken from GitHub [gists](https://gist.github.com). If you have the [**gistr**](https://cran.r-project.org/package=gistr) and [**fs**](https://fs.r-lib.org) packages installed, and have a GitHub account (PAT), you can deploy a `vegaspec` to gist, and it will appear as a block. If you have the [**magick**](https://github.com/ropensci/magick#readme) and [**webdriver**](https://github.com/rstudio/webdriver#README) packages (with PhantomJS installed), you can include preview and thumb images with the block. If a block (not necessarily yours) has a vegaspec, and the spec is in its own file, you can import a `vegaspec` from a block. 

TODO: Include caveats about git2r.

## Create a block

```{r}
library("vegawidget")
```

```{r eval=FALSE}
vw_create_block(spec_mtcars)
```

```{r echo=FALSE}
message(
  "block url: https://bl.ocks.org/06a512525fbe7a96e1a2028e1871b61c\n",
  " gist url: https://gist.github.com/06a512525fbe7a96e1a2028e1871b61c"
)
```

```{r eval=FALSE}
gistid <- vw_create_block_gistid(spec_mtcars)
```
```{r echo=FALSE}
message(
  "block url: https://bl.ocks.org/650e59f6a81cf63b9004ae291fe90667\n",
  " gist url: https://gist.github.com/650e59f6a81cf63b9004ae291fe90667"
)

gistid <- list(
  endpoint = "https://api.github.com",
  id = "650e59f6a81cf63b9004ae291fe90667"
)
```

## Retrieve a block

```{r}
vw_retrieve_block(gistid$id)
```
