---
title: "First Example"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r}
library("vegawidget")
```

In this vignette, we will go over a straightforward vegaspec: first as an overview, then examine each of its components, then highlight the mechanisms in this package we use to make the rendering. 

Here is our sample chart:

```{r echo=FALSE, comment = ""}
spec <- 
  list(
    `$schema` = "https://vega.github.io/schema/vega-lite/v2.json",
    description = "A simple bar chart with embedded data.",
    data = list(values = data_category),
    mark = "bar",
    encoding = list(
      x = list(field = "category", type = "ordinal"),
      y = list(field = "number", type = "quantitative")
    )
  ) %>%
  as_vegaspec()
```

```{r echo=FALSE}
spec
```

To see the JSON behind it, you can hover over the circle near the top-left of the chart and "Open in Vega Editor". As well, you can expand the section below: 

<details>
<summary>Open to see JSON</summary>
```{r echo=FALSE, comment = ""}
cat(format(vw_as_json(spec)))
```
</details>

----

We can build a vegaspec using lists; here's how we could build the sample chart:

```{r}
spec <- 
  list(
    `$schema` = vega_schema(),
    data = list(values = data_category),
    mark = "bar",
    encoding = list(
      x = list(field = "category", type = "ordinal"),
      y = list(field = "number", type = "quantitative")
    )
  ) %>%
  as_vegaspec()  
```

The first element, `` `$schema` ``,  helps the Vega-rendering system recognize what kind of specification this is: is it for Vega or Vega-Lite? We can use the function `vega_schema()`; its default-behavior is to return a string to indicate that we are using Vega-Lite.

The next element is used to specify the `data`. Although it may seem more-intuitive simply to specify this as a data-frame, [Vega-Lite also lets you specify](https://vega.github.io/vega-lite/docs/data.html) `data` using named-datasets, and URLs. We use the `values` element to signify that the data is provided in-line. The dataset, `data_category`, is included as a sample dataset with this package.

There is an element to specify the `mark`; in our case, each observation will appear as a `"bar"` on the chart. The available marks are detailed in the [Vega-Lite documentation](https://vega.github.io/vega-lite/docs/mark.html).

Finally, there is an element to specify each `encoding`. For each of the `x` and `y` encodings, the `field` and `type` are specified. The field is the name of the variable (column) in the `data`; the type is generally one of `"quantitative"`, `"nominal"`, `"ordinal"`, or `"temporal"`. Vega-Lite uses the `type` to use sensible defaults for axis- and label-formatting. As you might expect, more details are available at the [Vega-Lite documentation](https://vega.github.io/vega-lite/docs/encoding.html).

This list is piped to `as_vegaspec()`, which adds class-attributes to the `list`, letting us print and render in a much-less less-cumbersome way:

```{r}
spec
```

Behind the scenes, the print method does two things:

- it converts the list to JSON, using `vw_as_json()`
- it renders the JSON to an htmlwidget, using `vegawidget()`

This gives a very quick overview; for more details, please see the [vegaspec article](vegaspec.html) (focusing on the specification) and the [vegawidget article](vegawidget.html) (focusing on the rendering).

