---
title: "Create images"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

The main purpose of this article is to demonstrate use this package's image-generating functions. A secondary purpose is to help the package's developers confirm that these functions continue to work when things get changed. 

If you have **[nodejs](https://nodejs.org/en/)** installed, you can use these image-generating functions to convert vegaspecs (or vegawidgets) into SVG strings or write SVG files. With the **[rsvg](https://CRAN.R-project.org/package=rsvg)** package installed, you can get a bitmap array, or write PNG files.  

We have four image-generating functions; they each take either a vegaspec or a vegawidget as their main argument:

- `vw_to_svg()` returns a string for a SVG image
- `vw_to_bitmap()` returns a bitmap array that can be written to a PNG file
- `vw_write_svg()` writes an SVG file, invisibly returns the vegaspec or vegawidget
- `vw_write_png()` writes a PNG file, invisibly returns the vegaspec or vegawidget

As you can see, the function name indicates the action taken and the image-format.

In addition to a vegaspec or vegawidget, these are the principal arguments:

- `width`, and `height`, sent to `vw_autosize()` to specify the display dimensions. These refer to the *overall* dimensions of the image - not the dimensions of the plotting area.
- `scale`, to support display at retina resolutions. The default value is 1; for a retina image, you can specify a value of 2.
- `path`, for the image-writing functions, this indicates the local path to which the file is to be written.

As you are most likely already aware, there is a tradeoff between PNG and SVG. The PNG format is an image with finite resolution. The SVG format is a set of instructions on how to render an image, so it has no resolution limitations. In general, the size of an SVG file scales with the number of graphical marks it contains, whereas a PNG file scales with its pixel-width and pixel-height.


```{r, message=FALSE}
library("vegawidget")
library("htmltools")
library("glue")
```

```{r include=FALSE}
library("fs")
library("here")
library("conflicted")

# note to developers - the only way that this html file will 
# compile properly is to use pkgdown::build_site() or 
# pkgdown::build_article("image") - do not use the knit button

prefix_image <- "visual-regression"
dir_write <- here::here("docs", "reference", "figures")
dir_read <- here::here("reference", "figures")

path_write <- function(x) {
  fs::path(dir_write, glue::glue("{prefix_image}-{x}"))
}

path_read <- function(x) {
  fs::path(dir_read, glue::glue("{prefix_image}-{x}"))
}

# clean out old files
to_delete <- 
  dir_ls(
    dir_write, 
    regexp = glue::glue("^{fs::path(dir_write, prefix_image)}")
  )

path_rel(to_delete, start = here())
file_delete(to_delete)
```

## Reference 

For the purpose of comparing later renderings, our reference is this `mtcars` scatterplot. It is rendered, in HTML, with a plotting area of `width = 300` and `height = 300`.

```{r}
spec_mtcars
```

## Bitmap & PNG

To use the bitmap and PNG functions, you will need the **[rsvg](https://CRAN.R-project.org/package=rsvg)** package.

### To bitmap

The `vw_to_bitmap()` function returns a bitmap array. The package uses it primarly for converting the SVG to PNG; you may have other uses. 

We can show the effect of the `scale` argument by inspecting the dimensions of the bitmap array. Using the default, `scale = 1`:

```{r}
vw_to_bitmap(spec_mtcars) %>% dim()
```

At `scale = 2`:

```{r}
vw_to_bitmap(spec_mtcars, scale = 2) %>% dim()
```

### PNG output

The function `vw_write_png()` uses `vw_to_bitmap()` to write an image-file. In the examples that follow, we use a couple of internal helper-functions, `path_write()` and `path_read()`, to manage the file-locations within the **pkgdown** framework. For the purpose of this demonstration please disregard these functions, please interpret this only as a means to provide a `path`.

### Normal scale

The function `vw_write_png()` generates an image at 96 DPI. We will write out a PNG file, then display it at the same resolution.

**Expectation**: the image will be rendered at the same size as the [reference](#reference). 

```{r}
filename <- "mtcars-refrence.png"

vw_write_png(spec_mtcars, path = path_write(filename))

knitr::include_graphics(path_read(filename))
```

We note that the PNG appears slightly different from the HTML rendering. We suspect this is because the SVG-rendering engine, **[librsvg](https://github.com/GNOME/librsvg)**, works differently from your browser. In the future, we would like to find a way to make the rendering more-similar.

### Retina scale

By increasing the `scale` to 2, then decreasing the display-size by a factor of two, we can show this image at "retina" resolution. The code below is a little tortured, there should be an easier way. In the future, we would like for this to be the default method for knitting to a non-HTML medium (using PNG or SVG), and that it "just work".

```{r}
filename <- "mtcars-retina.png"
  
vw_write_png(spec_mtcars, path_write(filename), scale = 2)

# using knitr::include_graphics() with dpi not working, 
#  hacking a workaround

# determine the width of the image
meta <- png::readPNG(path_write(filename), info = TRUE)
width <- attr(meta, "info")[["dim"]][1]

# display width at half of image width
tags$img(src = path_read(filename), style = glue("width: {width / 2}px;"))
```

### Using vegawidget

This demonstrates that we can use `vw_write_png()` with vegawidgets. 

**Expectation**: this image will appear at the same size as the [reference](#reference).

```{r}
filename <- "mtcars-widget.png"
  
vw_write_png(vegawidget(spec_mtcars), path = path_write(filename))

knitr::include_graphics(path_read(filename))
```

## SVG

### To string

The `vw_to_svg()` function returns an HTML string:

```{r}
vw_to_svg(spec_mtcars)
```

This string can be rendered as standard HTML:

```{r}
HTML(vw_to_svg(spec_mtcars))
```

### Normal scale

**Expectation**: this image will appear identical to the [reference](#reference): 

```{r}
filename <- "mtcars-refrence.svg"

vw_write_svg(spec_mtcars, path = path_write(filename))

knitr::include_graphics(path_read(filename))
```

### Autosize

**Expectation**: We expect that the rendered image will have a *total* size of `width = 600` and `height = 300`, and to appear at retina resolution.

```{r}
filename <- "mtcars-wide.svg"

vw_write_svg(spec_mtcars, path = path_write(filename), width = 600, height = 300)

knitr::include_graphics(path_read(filename))
```

## Other visual-regressions

Although the rest of these do not use images, they are included in this article to make it easier to verify that things render as we expect.

### Autosize 

**Expectation**: the rendered spec will have a *total* size of `width = 600` and `height = 300`.

```{r}
vw_autosize(spec_mtcars, width = 600, height = 300)
```

**Expectation**: we will get the same result using the `vegawidget()` function:

```{r}
vegawidget(spec_mtcars, width = 600, height = 300)
```

**Expectation**: we will get the same result by setting `vega.width` and `vega.height` as knitr options, like so:

    ```{r vega.width=600, vega.height=300}
    spec_mtcars
    ```

```{r vega.width=600, vega.height=300}
spec_mtcars
```

### Vega

**Expectation**: this will be identical to the [reference](#reference) case, using the `vw_to_vega()` function to compile to a Vega specification:

```{r}
vw_to_vega(spec_mtcars)
```
