---
title: "Create images"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

The purpose of this article is to demonstrate use the image-generating functions, as well as to help the developers confirm that these functions continue to work when things get changed. If the [**webdriver**](https://github.com/rstudio/webdriver#README) package and PhantomJS are installed, you can use these image-generating functions to create or write images as PNG or SVG, using a vegaspec or vegawidget. 

We have four image-generating functions; they each take either a vegaspec or a vegawidget as their argument:

- `vw_to_png()` returns a data-URI string for a PNG image
- `vw_to_svg()` returns a string for a SVG image
- `vw_write_png()` writes a PNG file, invisibly returns the vegaspec or vegawidget
- `vw_write_svg()` writes an SVG file, invisibly returns the vegaspec or vegawidget

As you can see, the function name indicates the action taken and the image-format.

In addition to the vegaspec or vegawidget, these are the principal arguments:

- `scale`, i.e. scale-factor: ratio of the width (pixels) of the returned image to the width (pixes) of the rendered chart. If the `scale` is `1`, the resulting image may not look sharp on a retina display.

- `path`, for the image-writing functions, indicates the local path to which the file is to be written.

As you are most likely already aware, there is a tradeoff between PNG and SVG. The PNG format is an image with finite resolution. The SVG format is a set of instructions on how to render an image, so it has no resolution limitations. However, the size of an SVG file scales with the number of graphical marks it contains, whereas a PNG file scales with its pixel-width and pixel-height.

There is a problem using these functions on vegaspecs that load data from a URL; an [issue](https://github.com/vegawidget/vegawidget/issues/43) has been noted.

```{r, message=FALSE}
library("vegawidget")
library("glue")
```

```{r include=FALSE}
library("fs")
library("here")
library("conflicted")

# note to developers - the only way that this html file will 
# compile properly is to use pkgdown::build_site() or 
# pkgdown::build_article() - do not use the knit button

prefix_image <- "visual-regression"
dir_write <- here::here("docs", "reference", "figures")
dir_read <- here::here("reference", "figures")

path_write <- function(x) {
  fs::path(dir_write, glue::glue("{prefix_image}-{x}"))
}

path_read <- function(x) {
  fs::path(dir_read, glue::glue("{prefix_image}-{x}"))
}

# clean out old files
to_delete <- 
  dir_ls(
    dir_write, 
    regexp = glue::glue("^{fs::path(dir_write, prefix_image)}")
  )

path_rel(to_delete, start = here())
file_delete(to_delete)
```

## Reference 

Our reference is this `mtcars` scatterplot. We expect it to be rendered with a plotting area with `width = 300` and `height = 300`.

```{r}
spec_mtcars
```

## PNG

### To string

The `vw_to_png()` function returns a data-URI string that can be used as the `src` arrtibute of an `<img>` tag. Here's what such a string looks like:

```{r}
vw_to_png(spec_mtcars)
```

Here's how it can be used in a tag to render the string as an image:

```{r}
library("htmltools")

tags$img(src = vw_to_png(spec_mtcars))
```

The rest of the PNG demonstration focuses on the image-writing function, `vw_write_png()`, which itself uses `vw_to_png()`. In the examples that follow, we use a couple of internal helper-functions, `path_write()` and `path_read()`, to manage the file-locations within the **pkgdown** framework. For the purpose of this demonstration please disregard these functions, focusing instead on that we are providing a `path`.

### Normal scale

The function `vw_write_png()` generates an image at 96 DPI. We will write out a PNG file, then display it at the same resolution.

**Expectation**: the image will be rendered at the same size as the [reference](#reference). However, we not using retina resolution so it will appear a bit fuzzy on non-retina monitors.

```{r}
filename <- "mtcars-refrence.png"

vw_write_png(spec_mtcars, path = path_write(filename))

knitr::include_graphics(path_read(filename))
```

### Retina scale

To get retina resolution, we use the `scale` argument in `vw_write_png()`, then adjust the `dpi` argument in `include_graphics()`. 

**Expectation**: the image will appear indentical to the [reference](#reference) at retina resolution. 

```{r}
filename <- "mtcars-retina.png"
  
vw_write_png(spec_mtcars, path_write(filename), scale = 2)

# using knitr::include_graphics() with dpi not working, 
#  hacking a workaround

# determine the width of the image
meta <- png::readPNG(path_write(filename), info = TRUE)
width <- attr(meta, "info")[["dim"]][1]

# display width at half of image width
tags$img(src = path_read(filename), style = glue("width: {width / 2}px;"))
```

### Autosizing at normal scale

The image functions can autosize a chart to a given `width` and `height`.

**Expectation**: the rendered image will have a *total* size of `width = 600` and `height = 300`, but at non-retina resolution:

```{r}
filename <- "mtcars-wide.png"
  
vw_write_png(spec_mtcars, path_write(filename), width = 600, height = 300)

knitr::include_graphics(path_read(filename))
```

### Using vegawidget

This demonstrates that we can expect use `vw_write_png()` with vegawidgets. 

**Expectation**: this image will appear at the same size as the [reference](#reference), but at non-retina resolution.

```{r}
filename <- "mtcars-widget.png"
  
vw_write_png(vegawidget(spec_mtcars), path_write(filename))

knitr::include_graphics(path_read(filename))
```

## SVG

### To string

The `vw_to_svg()` function returns an HTML string:

```{r}
vw_to_svg(spec_mtcars)
```

This string can be rendered as standard HTML:

```{r}
HTML(vw_to_svg(spec_mtcars))
```

### Normal scale

**Expectation**: this image will appear identical to the [reference](#reference): 

```{r}
filename <- "mtcars-refrence.svg"

vw_write_svg(spec_mtcars, path_write(filename))

knitr::include_graphics(path_read(filename))
```

### Autosize

**Expectation**: We expect that the rendered image will have a *total* size of `width = 600` and `height = 300`, and to appear at retina resolution.

```{r}
filename <- "mtcars-wide.svg"

vw_write_svg(spec_mtcars, path_write(filename), width = 600, height = 300)

knitr::include_graphics(path_read(filename))
```

### Retina scale

The `scale` argument in `vw_write_svg()` may not be useful becuse the SVG itself is scalable. 

**Expectation**: this will appear at double-size to [reference](#reference), with retina resolution: 

```{r}
filename <- "mtcars-retina.svg"

vw_write_svg(spec_mtcars, path_write(filename), scale = 2)

knitr::include_graphics(path_read(filename))
```

### Vegawidget

As above, this demonstrates that we can expect use `vw_write_svg()` with vegawidgets. 

**Expectation**: this image will appear at the same size as the [reference](#reference), at non-retina resolution.

```{r}
filename <- "mtcars-widget.svg"

vw_write_svg(vegawidget(spec_mtcars), path_write(filename))

knitr::include_graphics(path_read(filename))
```

## Other visual-regressions

Although the rest of these do not use images, they are included in this article to make it easier to verify that things render as we expect.

### Autosize 

Next, we test using `vw_autosize()`. 

**Expectation**: the rendered spec will have a *total* size of `width = 600` and `height = 300`.

```{r}
vw_autosize(spec_mtcars, width = 600, height = 300)
```

**Expectation**: we will get the same result using the `vegawidget()` function:

```{r}
vegawidget(spec_mtcars, width = 600, height = 300)
```

**Expectation**: we will get the same result by setting `vega.width` and `vega.height` as knitr options, like so:

    ```{r vega.width=600, vega.height=300}
    spec_mtcars
    ```

```{r vega.width=600, vega.height=300}
spec_mtcars
```

### Vega

**Expectation**: this will be identical to the [reference](#reference) case, using the `vw_to_vega()` function to compile to a Vega specification:

```{r}
vw_to_vega(spec_mtcars)
```
