---
title: "Import into Other Packages"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

The principal motivation for the **vegawidget** package is to provide other packages that build Vega or Vega-Lite specifications a resuable set of functions.

To make this truly resuable... `use_pipe()`...

Your package might have an S3 class. For *altair*, this is `altair.vegalite.v2.api.TopLevelMixin`.

Call `use_vegawidget()`, it will:

- ...

You need to adapt one thing, then everything else "just works".

```{r s3}
spec_mtcars_s3 <- structure(
  spec_mtcars,
  class = c("my_class_name")
)

class(spec_mtcars_s3)
```

To take full advantage of these rendering functions, we have to define a method, `as_vegaspec()` for this class, and also define `print()` and `knit_print()` methods.

```{r methods}
as_vegaspec.my_class_name <- function(spec, ...) {
  
  # reverts to list
  spec <- unclass(spec)

  # as_vegaspec has a list method
  vegawidget::as_vegaspec(spec, ...)
}

print.my_class_name <- function(x, ...) {
  
  x <- vegawidget::as_vegaspec(x)
  
  print(x, ...)
}

knit_print.my_class_name <- function(x, ..., options = NULL){
  
  x <- vegawidget::as_vegaspec(x)

  knitr::knit_print(x, ..., options = options)
}
```

With these defined, you can render your specification by printing:

```{r}
spec_mtcars_s3
```
