---
title: "Render using vegawidget"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

The central *side-effect* of this package is to render a *vegawidget*, which is also an htmlwidget. A vegawidget is built by the `vegawidget()` function, using a `vegaspec`. 

The `vegawidget()` function lets you call is provided for [vega-embed](https://github.com/vega/vega-embed) options. These options can be specified as an argument to the `vegawidget()` function, or as an R option, `vega.embed`. The R options `vega,embed`, `vega.height`, and `vega.width` can be useful when printing or creating an image from a spec. 

If the [**webshot**](https://CRAN.R-project.org/package=webshot) package and PhantomJS are installed, knitting to a non-html format should "just work".

## Using vegawidget arguments

When you print a `vegaspec`, its print method calls `vegawidget()`.

```{r}
library("vegawidget")

vegawidget(spec_mtcars) 
```
 
 In addition to `spec`, the function `vegawidget()` has arguments to `embed` and to specify the overall dimensions, `width` and `height`. 

The `embed` argument is used to specify the embedding options, according to the API to the [vega-embed library](https://github.com/vega/vega-embed#options). There are a lot of options; to make things easier, you can use the helper function `vega_embed()`, it mirrors the options available to the version of the vega-embed library supported by this package (see `vega_version()`). The most important arguments are:

- `renderer`: `character`, either `"canvas"` (default) or `"svg"`
- `actions`: named `list` of logicals to indicate the presence of action-links; names (default value) can be `export` (`TRUE`), `source` (`TRUE`), `compiled` (`FALSE`), or `editor` (`TRUE`).

Here are some examples of various options:

```{r}
vegawidget(spec_mtcars, embed = vega_embed(renderer = "svg"))
```

```{r}
vegawidget(spec_mtcars, embed = vega_embed(actions = FALSE))
```

The arguments `width` and `height` in `vegawidget()` are used to autosize the `spec` before rendering, using `vw_autosize()`.

```{r}
vegawidget(spec_mtcars, width = 300, height = 300)
```

Here, the `width` and `height` refer to the overall dimensions of the rendered vegaspec.
Please see the [vegaspec article](vegaspec.html) for more details on how sizing works.

## Using knitr chunk-options

In the previous section, you saw how to specify the rendering options using the `vegawidget()` function explicitly. You may wish to set the options implicitly: you can do so using the R options `vega.embed`, `vega.width`, and `vega.height`.

One way to set these options is in a knitr chunk:

    ```{r vega.embed=vegawidget::vega_embed(actions = FALSE)}
    spec_mtcars
    ```

```{r vega.embed=vegawidget::vega_embed(actions = FALSE), echo=FALSE}
spec_mtcars
```

    ```{r vega.width=300, vega.height=300}
    spec_mtcars
    ```

```{r vega.width=300, vega.height=300, echo=FALSE}
spec_mtcars
```

## Extending using JavaScript

Here, we recreate this Vega-Lite [streaming tutorial](https://vega.github.io/vega-lite/tutorials/streaming.html).

```{r}
spec_table <-
  list(
    `$schema` = vega_schema(),
    width = 400,
    data = list(name = "table"),
    mark = "line",
    encoding = list(
      x = list(
        field = "x", 
        type = "quantitative",
        scale = list(zero =  FALSE)
      ),
      y = list(field = "y", type = "quantitative"),
      color = list(field = "category", type = "nominal")
    )
  ) %>%
  as_vegaspec()
```

The `elementId` argument is passed on to `htmlwidget::createWidget()`. 

```{r}
vegawidget(spec_table, elementId = "streaming-demo")
```

```{js}
HTMLWidgets.addPostRenderHandler(function() {

  getVegaView('#streaming-demo')
    .then(function(view) {
      /**
       * Generates a new tuple with random walk.
       */
      function newGenerator() {
          var counter = -1;
          var previousY = [5, 5, 5, 5];
          return function () {
              counter++;
              var newVals = previousY.map(function (v, c) { return ({
                  x: counter,
                  y: v + Math.round(Math.random() * 10 - c * 3),
                  category: c
              }); });
              previousY = newVals.map(function (v) { return v.y; });
              return newVals;
          };
      }
  
      var valueGenerator = newGenerator();
      var minimumX = -100;
      window.setInterval(function () {
          minimumX++;
          var changeSet = vega.changeset().insert(valueGenerator()).remove(function (t) { return t.x < minimumX; });
          view.change('table', changeSet).run();
      }, 1000);  
    });

});
```




