---
title: "Specify using vegaspec"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

The central object of this package is a vegaspec, short for "Vega specification". This can refer to either a Vega or Vega-Lite specification; in our examples, we work almost exclusively with Vega-Lite.

This article will showcase functions that:

- build a vegaspec: `as_vegaspec()`
- examine a vegaspec: `vw_examine()`
- print vegaspecs: `print()` and `knit_print()` methods
- convert a vegaspec to JSON: `vw_as_json()`
- resize a vegaspec: `vw_autosize()`
- complie a Vega-Lite spec to Vega: `vw_to_vega()`

The article ends with some "gotchas" (please contribute more as you find them).

## Vegaspec

A vegaspec is created using the constructor `as_vegaspec()`. You can coerce to a `vegaspec` object from a `list`, `character` or a **jsonlite** `json` object. 

Vega and Vega-Lite expect JSON. However, within R, it is easier to work with lists. This is one of the philosophies of this package: a vegaspec "lives" as a list, and then is corerced to JSON when rendered as a vegawidget.

```{r}
library("vegawidget")

list_anscombe_1 <- 
  list(
    `$schema` = "https://vega.github.io/schema/vega-lite/v2.json",
    data = list(values = anscombe),
    mark = "point",
    encoding = list(
      x = list(field = "x1", type = "quantitative"),
      y = list(field = "y1", type = "quantitative")
    )
  ) %>%
  as_vegaspec()
```

Accordingly, the `vegaspec` S3 class inherits from `list`.

```{r}
str(spec_anscombe_1)
```

A couple of points before going further. We have a helper function, `vega_schema()`, that you can use to generate the string for the `$schema` element. This can be useful if you are writing a vegaspec manually. We also have a function, `vw_examine()`, that you can use to, well, examine a vegaspec.

```{r}
list_anscombe_1 <- list(
  `$schema` = vega_schema(),  # defaults to major version of "vega_lite"
  data = list(values = anscombe),
  mark = "point",
  encoding = list(
    x = list(field = "x1", type = "quantitative"),
    y = list(field = "y1", type = "quantitative")
  )
) 

spec_anscombe_1 <- as_vegaspec(list_anscombe_1)

vw_examine(spec_anscombe_1)
```

The `vw_examine()` function is a thin wrapper around `listviewer::jsonedit()`; it can be handy to see what is "going on" inside a vegaspec.
 
The `vega_schema()` function has an argument to specify the `library` as either `"vega-lite"` or `"vega"`, and another that indicates to use the `major` version of the `library`. It has a complematary function, `vw_spec_version()`, that can be used to return the `library` and `version` of the schema of a vegaspes:

```{r}
vw_spec_version(spec_anscombe_1)
```

## Printing

A few words about `print()` and `knit_print()` methods, then refer to the vegawidget article for more details.  

## JSON

When a vegaspec is rendered as a vegawidget, the spec is coerced automatically to JSON by the `vegawidget()` function. Like other JSON-based systems, Vega and Vega-Lite have their own particularities in their JSON format. These are taken into account by the function `vw_as_json()`, which is a wrapper around `jsonlite::toJSON()`. You may find this handy if you want to write a vegaspec to disk or elsewhere.

We can examine the JSON:

```{r}
json_anscombe_1 <- vw_as_json(spec_anscombe_1)

vw_examine(json_anscombe_1)
```

You will notice a difference between the original, `spec_anscombe_1`, and the json version, `json_ancombe_1`: the `data$values` element has been transposed. As you know, an R data frame is column-oriented; Vega expects data to be row-oriented. Our function `vw_as_json()` takes care of this so that the spec will render properly. This is a one-way trip; it is an exercise for the reader to show that `as_vegaspec()` does not change this back.

## Sizing

Sizing can be a tricky issue in Vega-Lite. 

The first thing to keep in mind is that the default interpretation of width and height refers to the dimensions of the plotting area. For example, we can set `width` and `height` of our Anscombe spec:

```{r}
spec_anscombe_size <- spec_anscombe_1

spec_anscombe_size$width  <- 300
spec_anscombe_size$height <- 300

spec_anscombe_size
```

Here, the plotting area is 300 $\times$ 300. If we want to set the dimension of the entire plot, we can use the `vw_autosize()` function:

```{r}
spec_anscombe_autosize <- 
  vw_autosize(spec_anscombe_size, width = 300, height = 300)

spec_anscombe_autosize
```

In this context, the `width` and `height` are defined as the dimensions of the entire plot including axes, labels, padding, etc.

There are limitations to autosizing; the most prominent is that specifying width and height is [effective only for single-view and layered specifications](https://vega.github.io/vega-lite/docs/size.html#limitations). It will not work for specifications with multiple views (e.g. `hconcat`, `vconcat`, `facet`, `` `repeat` ``); this will issue a warning that there will be no effect on the specification when rendered.

For more information, see the Vega-Lite documentation on [sizing](https://vega.github.io/vega-lite/docs/size.html#limitations).

## Compile to Vega

When rendering a Vega-Lite specification, the vega-lite JavaScript library will compile the Vega-Lite specification into Vega. If you have the **V8** package installed, and want to compile a Vega-Lite specicifation to Vega, for your own purposes, you can use the `vw_to_vega()` function.

```{r}
spec_anscombe_1_vega <- vw_to_vega(spec_anscombe_1)

spec_anscombe_1_vega
```

```{r}
vw_spec_version(spec_anscombe_1_vega)
```

```{r}
vw_examine(spec_anscombe_1_vega)
```

You can autosize a Vega specification.

```{r}
vw_autosize(spec_anscombe_1_vega, width = 300, height = 300)
```

## Gotchas

There are a few things to keep your eyes open-for if you are building vegaspec: 

- If the JSON needs a `null`, supply a `NULL` from R.

- If the JSON needs an element called `repeat`, keep in mind that this is a reserved word in R. You can specify such an element using backticks, e.g. ``list(`repeat` =  ...)``.



