---
title: "Specify using vegaspec"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

The central object of this package is a *vegaspec*, short for "Vega specification". This can refer to either a Vega or Vega-Lite specification; in our examples, we work almost exclusively with Vega-Lite.

This article will showcase functions that:

- build a vegaspec: `as_vegaspec()`
- examine a vegaspec: `vw_examine()`
- convert a vegaspec to JSON: `vw_as_json()`
- resize a vegaspec: `vw_autosize()`
- compile a Vega-Lite spec to Vega: `vw_to_vega()`

The article ends with some "gotchas" (please contribute more as you find them).

## Building

A vegaspec is created using the constructor `as_vegaspec()`, which takes a `list`, `character` or a **jsonlite** `json` object. 

Vega and Vega-Lite expect JSON. However, within R, it is easier to work with lists. This is one of the underpinnings of this package: a vegaspec "lives" as a list, and then is coerced to JSON when rendered as a vegawidget.

Let's start with a bar-chart example:

```{r}
library("vegawidget")

spec_category <- 
  list(
    `$schema` = vega_schema(),
    data = list(values = data_category),
    mark = "bar",
    encoding = list(
      x = list(field = "category", type = "nominal"),
      y = list(field = "number", type = "quantitative")
    )
  ) %>%
  as_vegaspec()

spec_category
```

Looking at the structure of `spec_category`, we see that the `vegaspec` S3 class inherits from `list`.

```{r}
str(spec_category)
```

Note the helper function, `vega_schema()`, that you can use to generate the string for the `$schema` element. This can be useful if you are writing a vegaspec manually. 


## Examining

We also have a function, `vw_examine()`, that you can use to, well, examine a vegaspec.

```{r}
vw_examine(spec_category)
```

You might note that, here, the `data$values` element of the vegaspec has a column-major format. In R, data frames are lists of vectors; each of the columns is a vector. In the Vega framework, data sets have a row-major format. What we see above is the specification *before* it is rendered. In the rendering process, data-frames are converted to row-major data. 

## Converting to JSON

When a vegaspec is rendered as a vegawidget, the spec is coerced automatically to JSON by the `vegawidget()` function. Like other JSON-based systems, Vega and Vega-Lite have their own particularities in their JSON format. These are taken into account by the function `vw_as_json()`, which is a wrapper around `jsonlite::toJSON()`. You may find this handy if you want to write a vegaspec to disk or elsewhere.

Let's examine the JSON for our example code:

```{r}
spec_category %>% 
  vw_as_json() %>% 
  vw_examine()
```
Here, the `data$values` element has been transposed. The function `vw_as_json()` takes care of this so that the spec will render properly. 

## Resizing

Sizing can be a tricky issue in Vega-Lite. 

The first thing to keep in mind is that the default interpretation of width and height refers to the dimensions of the plotting area. For example, we can set `width` and `height` of our example spec:

```{r}
spec_category_size <- spec_category

spec_category_size$width  <- 300
spec_category_size$height <- 300

spec_category_size
```

Here, the plotting area is 300 $\times$ 300. If we want to set the dimension of the entire plot, we can use the `vw_autosize()` function:

```{r}
spec_category_autosize <- 
  spec_category %>%
  vw_autosize(width = 300, height = 300)

spec_category_autosize
```

In this context, the `width` and `height` are defined as the dimensions of the entire plot including axes, labels, padding, etc.

There are limitations to autosizing; the most prominent is that specifying width and height is [effective only for single-view and layered specifications](https://vega.github.io/vega-lite/docs/size.html#limitations). It will not work for specifications with multiple views (e.g. `hconcat`, `vconcat`, `facet`, `` `repeat` ``); this will issue a warning that there will be no effect on the specification when rendered.

For more information, see the Vega-Lite documentation on [sizing](https://vega.github.io/vega-lite/docs/size.html#limitations).

## Compile to Vega

When rendering a Vega-Lite specification, the vega-lite JavaScript library will compile the Vega-Lite specification into Vega. If you have the **V8** package installed, and want to compile a Vega-Lite specification to Vega, for your own purposes, you can use the `vw_to_vega()` function.

To be clear, if you are using only Vega-Lite, this is not something you should ever need to do. This may be useful if you need to design a specification using Vega, and you want get things started using a Vega-Lite specification.

```{r}
spec_category_vega <- vw_to_vega(spec_category)

spec_category_vega
```

We can use the `vw_spec_version()` function to check the version of the specification:

```{r}
vw_spec_version(spec_category_vega)
```

You can examine the result to see how much-more verbose is the Vega specification, compared to the Vega-Lite specification.

```{r}
vw_examine(spec_category_vega)
```

## Gotchas

There are a few things to keep your eyes open-for if you are building vegaspec: 

- If the JSON needs a `null`, supply a `NULL` from R.

- If the JSON needs an element called `repeat`, keep in mind that this is a reserved word in R. You can specify such an element using backticks, e.g. ``list(`repeat` =  ...)``.



