---
title: "Visual Regression"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

The purpose of this document is to use our eyeballs to vaildate the rendering.

```{r}
library("magrittr")
library("vegawidget")

options(
  vega.embed = vegawidget::vega_embed(defaultStyle = TRUE)
)
```

## Reference 

Our first test is the standard mtcars scatterplot, where the plotting area is 300 * 300.

```{r}
spec_mtcars
```

```{r}
print(spec_mtcars)
```

```{r}
print(2)
```

## Vega-Lite 

Let's convert this spec to JSON, then back to a vegaspec, then render it:

```{r}
spec_mtcars %>%
  as_json() %>%
  as_vegaspec()
```

Next, we test using `autosize()`. The size of the div will be 300 * 300; the plotting area will be smaller.

```{r}
autosize(spec_mtcars, width = 600, height = 300)
```

We can do the same thing using the `vegawidget()` function:

```{r}
vegawidget(spec_mtcars, width = 600, height = 300)
```

Or by setting `vega.width` and `vega.height` as knitr options, like so:

    ```{r vega.width=600, vega.height=300}
    spec_mtcars
    ```

```{r vega.width=600, vega.height=300}
spec_mtcars
```

## Vega

There should no difference from the first case when we use the `to_vega()` function to compile to a Vega specification, then render:

```{r}
spec_vega_mtcars <- to_vega(spec_mtcars) %>% print()
```

Should only need to confirm in one way that autosize works for Vega specs.

```{r}
autosize(spec_vega_mtcars, width = 600, height = 300)
```

## PNG

We should see this at regular size, then at expanded size:

<img src=`r to_png(spec_mtcars, scale = 2)`>



