---
title: "Render using vegawidget"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

The central *side-effect* of this package is to render a *vegawidget*, an htmlwidget. 

A vegawidget is built by the `vegawidget()` function, using a `vegaspec`. However, in many cases, your call to `vegawidget()` may be implied in a `print` method, rather than being made explicitly. 

The `vegawidget()` function has three main arguments:

- `embed` - to provide [vega-embed](https://github.com/vega/vega-embed)
- `width`, `height` - to provide width and height to `vw_autosize()`

Because your call to vegawidget may be implicit, there are a number of ways to provide these arguments.

- using explicit arguments (`embed`, `width`, `height`) to the `vegawidget()` function 
- setting R options (`vega.embed`, `vega.width`, and `vega.height`) in a knitr chunk
- setting R options using `options()`

These methods are listed in order of precedence; the `embed` argument to the `vegawidget()` function will be used in favor of `getOption("vega.embed")`. Furthermore, if you specify the `embed` argument, `vegawidget()` will ignore any part of the embed-object that may have been set in `getOption("vega.embed")`.

These options make the most sense if you are rendering an HTML document. If you are rendering to a non-HTML format, this package's image-writing functions will conspire to try to do the right thing.

## Using vegawidget arguments

When you print a `vegaspec`, its print method calls `vegawidget()`.

```{r}
library("vegawidget")

spec_mtcars 
```
 
In addition to `spec`, the function `vegawidget()` has arguments to `embed` and to specify the overall dimensions, `width` and `height`. Setting the options using `vegawidget()` arguments will override options set any other way.

The `embed` argument is used to specify the embedding options, according to the API to the [vega-embed library](https://github.com/vega/vega-embed#options). There are a lot of options; to make things easier, you can use the helper function `vega_embed()`, it mirrors the options available to the version of the vega-embed library supported by this package (see `vega_version()`). The most important arguments are:

- `renderer`: `character`, either `"canvas"` (default) or `"svg"`
- `actions`: named `list` of logicals to indicate the presence of action-links; names (default value) can be `export` (`TRUE`), `source` (`TRUE`), `compiled` (`FALSE`), or `editor` (`TRUE`).

For example, you can change the renderer to SVG:

```{r}
vegawidget(spec_mtcars, embed = vega_embed(renderer = "svg"))
```

You can remove the actions "bug" that is otherwise placed at the top-right of the chart:

```{r}
vegawidget(spec_mtcars, embed = vega_embed(actions = FALSE))
```

The arguments `width` and `height` in `vegawidget()` are used to autosize the `spec` before rendering, using `vw_autosize()`:

```{r}
vegawidget(spec_mtcars, width = 300, height = 300)
```

Here, the `width` and `height` refer to the overall dimensions of the rendered vegaspec. Please see the [vegaspec article](vegaspec.html) for more details on how sizing works.

The `base_url` argument lets you specify the a common root for when you specify data using a URL. Consider this vegaspec:

```{r}
spec_precip <-
  list(
    `$schema` = vega_schema(),
    data = list(url = "seattle-weather.csv"),
    mark = "tick",
    encoding = list(
     x = list(field = "precipitation", type = "quantitative")
    )
  ) %>%
  as_vegaspec()
```

We have included this file as an example dataset for this package; it is also available as a part of **[vega-datasets](https://github.com/vega/vega-datasets)**

```{r}
path_local <- system.file("example-data", package = "vegawidget")
url_remote <- "https://vega.github.io/vega-datasets/data"
```

By using a `base_url`, you can specify the file-location at rendering, rather than as a part of the specification:

```{r, eval=FALSE}
# not run
vegawidget(spec_precip, base_url = path_local)
```

Using a local path will not work properly in this article because you do not have the same path on your computer. Local paths should be used only for local work; this is a variation on Jenny Bryan's famous [safety-tip](https://twitter.com/hadleywickham/status/940021008764846080).

If you are sharing your work, using a remote URL will be more robust, but requires that the data be publicly accessible.

```{r}
vegawidget(spec_precip, base_url = url_remote)
```

## Setting R options

In the previous section, you saw how to specify the rendering options using the `vegawidget()` function explicitly. You may wish to set the options implicitly: you can do so using the R options `vega.embed`, `vega.width`, and `vega.height`.

One way to set these options is in a knitr chunk:

    ```{r vega.embed=vegawidget::vega_embed(actions = FALSE)}
    spec_mtcars
    ```

```{r vega.embed=vegawidget::vega_embed(actions = FALSE), echo=FALSE}
spec_mtcars
```

    ```{r vega.width=300, vega.height=300}
    spec_mtcars
    ```

```{r vega.width=300, vega.height=300, echo=FALSE}
spec_mtcars
```

We can set options explicitly using the `options()` function. There are a few ways of the ways you can do this:

- put an `options()` call in your `.Rprofile`
- put an `options()` call in a code chunk in your `.Rmd` file
- use the [**config**](https://cran.r-project.org/package=config) package, with `!expr options(...)` in your YML configuration file.

Here, we use a variation of the second option, making an explicit `options()` call to set the width and height:

```{r}
# render spec using temporary options
withr::with_options(
  new = list(vega.width = 300, vega.height = 300),
  vegawidget(spec_mtcars)
)
```

## Knitting strategies

There are two things to keep in mind when knitting a document that contains a vegawidget:

- the type of document: HTML-based format or non-HTML-based format.
- the sizing of the vegawidget: different formats have different sizing options available.

### Format

In some situations, you may knit to an HTML-based format, like `html_document`; in others you may knit to a non-HTML-based format, like `github_document`, or `pdf_document`. This article is an example of knitting to an `html_document`; this package's [GitHub README](https://github.com/vegawidget/vegawidget/blob/master/README.md) is an example of knitting to a `github_document` (using SVG).

If you are knitting to an HTML-based format, the only supported options are `vega.width`, `vega.height` (as pixels) and `vega.embed` options (as a list). 

If you are knitting to a **non**-HTML-based format, you additionally have the options `dev` (`"png"`, `"svg"`, or `"pdf"`), `out.width` and `out.height`; sizing is discussed in the following section. 

Knitting to a non-HTML-based format calls this package's [image](image.html) functions. This requires:

- **[nodejs](https://nodejs.org)** to be installed on your computer.
- the packages **[rsvg](https://CRAN.R-project.org/package=rsvg)** and **[png](https://CRAN.R-project.org/package=png)**.
- if you are using MacOS, knitr's SVG device will require an X11 system, such as [XQuartz](https://www.xquartz.org), to be installed.

 If you are knitting to a LaTeX format (e.g. `pdf_document`) and you specify `dev` as `"svg"`, it will be implemented as `"pdf"`.
  
### Sizing

The biggest thing to keep in mind about sizing a Vega visualization is that very often the chart tells you how much space it needs, rather than than you tell it how much space it has available. In the future, it may reveal itself how to manage better this "conversation" between you and the chart. 

One consequence of this "Vega-tells-you" sizing philosophy is that the options `fig.width` and `fig.height` are not useful. Either the chart will tell you the size based on its defaults, or you will specify some dimensions in the vegaspec: `width`, `height`, or size of rectangles. If your chart is a single view, i.e. not faceted, concatenated, or repeated, you can specify the size of the chart using the options `vega.width` and `vega.height`.

  - `vega.width` and `vega.height` are passed to `vegawidget()` as `width` and `height`, respectively. These values are coerced to numeric, so it is ineffective to specify a percentage. They are passed to `vw_autosize()`, which will resize the chart, if [possible](https://vega.github.io/vega-lite/docs/size.html#limitations).

  - For HTML-formats, setting the values of `out.width` and `out.height` will not affect the rendered size of your chart.
  
  - For non-HTML-formats, you can use  `out.width` or `out.height` to scale the size of the image within your document.
Because the image will already have an aspect ratio, it is recommended to specify no more than one of these.

