---
title: "Extend using Shiny"
date: "`r Sys.Date()`"
output: rmarkdown::html_document
---

```{r}
library("vegawidget")
```

The purpose of this article is to show how you can use Shiny to extend Vega's capabilities. In these examples, we use the [vega-view](https://github.com/vega/vega/tree/master/packages/vega-view) API to interact with a Vega chart using Shiny We explore similar ideas in an article on [extending using JavaScript](javascript.html).

Because this article is an HTML document (as opposed to a Shiny app), we cannot show the app directly in this document. However, this package offers a function `vw_shiny_demo()` that, when called without arguments, lists all of the demonstration apps included with this package; to run a particular app, simply use its name as the argument:

```r
vw_shiny_demo("data-set-get")
```

Once a Vega chart is built, there are three things you can access through Shiny:

- [events](https://vega.github.io/vega/docs/api/view/#event-handling): using `vw_shiny_get_event()`
- [signals](https://vega.github.io/vega/docs/api/view/#signals): using `vw_shiny_get_signal()` and `vw_shiny_set_signal()`
- [data](https://vega.github.io/vega/docs/api/view/#data-and-scales): using `vw_shiny_get_data()` and `vw_shiny_set_data()`

As we go through the examples, please keep in mind:

- the `vw_shiny_get_*()` functions **return reactive expressions**; these reactives return values.
- the `vw_shiny_set_*()` functions **act as Shiny observers**; these observers carry out side-effects.

## Getting events {#events}

This is a shortened version of the [events section](javascript.html#events) in the [JavaScript article](javascript.Rmd); the concepts are the same, but the implementation is far simpler using Shiny. 

Event streams are [used extensively within Vega](https://vega.github.io/vega/docs/event-streams), enabling interactivity. By [adding event-listeners](https://vega.github.io/vega/docs/api/view/#view_addEventListener) to the `view`, we can specify actions to be taken in response to a particular type of event, such as `"mouseover"`, `"click"`, `"keypress"`, etc.

The Shiny demo for events features a scatterplot. When a mark on the chart is clicked, it returns to Shiny the data observation (datum) associated with that mark.

You can see this in action by running:

```r
vw_shiny_demo("event-get")
```

The operative function in the `server()` function is `vw_shiny_get_event()`:

```r
# the event returns the backing-data as a list
rct_list_click <-
  vw_shiny_get_event("chart", event = "click", body_value = "datum")
```

The function is called with three arguments:

- `outputId`, the scatterplot has the outputId `"chart"`
- `event`, we want to respond to `"click"` events on the scatterplot
- `body_value`, please see the following text...

The vega-view API requires that you supply a handler-function when you [add an event listener](https://vega.github.io/vega/docs/api/view/#view_addEventListener). Inconveniently for R users, this handler-function has to be JavaScript. This package does a few things to make working with JavaScript handlers a little easier.

An event-handler function is expected to have two arguments, `event` instance and the scenegraph `item`. Because all possible event-handlers will have the same arguments, we need specify only the *body* of the function in `vw_shiny_get_event()`.

Instead of specifying the entire body of a JavaScript function, this package provides a library of bodies of (what we think are) commonly-used handler functions. To see the available event-handlers, call `vw_handler_event()` without arguments:

```{r}
vw_handler_event()
```

We wish to return the data-observation behind the mark associated with the event, so we can specify to use `"datum"` handler. The `vw_shiny_get_event()` function knows to check the library first, so you can either supply the entire body of a JavaScript handler-function, or you can supply the name of a handler-function in the library.

As such, the following calls are equivalent:

```r
# name of a body in the event-handler library
vw_shiny_get_event(..., body_value = "item") 

# body of a JavaScript function
vw_shiny_get_event(..., body_value = "return item;") 
```

The code supplied for `body_value` need only return a *value* - the `vw_shiny_get_event()` function will take care of composing this with additional code that will make that *value* available to Shiny.

Finally, the `vw_shiny_get_event()` function returns a reactive expression that evaluates to the latest *value* returned from the event-handler.

To see the same logic implemented in pure JavaScript, see the [events demo](javascript.html#demo-events) of the JavaScript article.

## Setting and getting signals

As with the events section, this is a shortened version of the [signals section](javascript.html#signals) in the [JavaScript article](JavaScript).

Signals are a [formal part of the Vega specification](https://vega.github.io/vega/docs/signals), but not officially a part of the Vega-Lite specification. They are "dynamic variables that parameterize a visualization and can drive interactive behaviors. Signals can be used throughout a Vega specification, for example to define a mark property or data transform parameter."

Even though signals are not defined in Vega-Lite, this does not mean that we cannot use them. Before being rendered, Vega-Lite specifications are compiled into Vega specifications. Once rendered, we can interact with the Vega signals, as long as we know what they are named.

For example, signals are used to drive the behavior of Vega-Lite [selections](https://vega.github.io/vega-lite/docs/selection.html). As of the start of 2019, the Vega-Lite development team are [discussing](https://github.com/vega/vega-lite/issues/3338) if, and how, signals might be introduced to Vega-Lite.

One way we can "hack" our way to a signal is to use the signal the Vega-Lite specification, then *define* the signal by patching the compiled Vega specification. Vega-Lite developer Dominik Moritz demonstrates this technique in an [Observable notebook](https://beta.observablehq.com/@domoritz/rotating-earth). 

Here, we will use the same technique to create an interactive histogram. As you know, whenever you create a histogram it is a good idea to experiment with different bin-widths, to give yourself a chance to see all of the interesting structure in the data. We will use the `data_seattle_hourly` dataset, included with this package, to look at the distribution of temperatures in Seattle in 2010.

You can see this in action by running:

```r
vw_shiny_demo("signal-set-get")
```

Using the display mode of the app you can see the R code used to build it. In the Vega-Lite spec, in the `x` `encoding`, we have defined the `bin` `step` as a `signal` named `"bin_width"`. This is not legal in Vega-Lite.
To make this work, we have to patch the compiled Vega spec with a definition of the signal. We can include the patch as an option to `vega_embed()`.

You also see in the code that we transform the value returned by the input, which varies between -1 and 1, to the bin-width which varies between 0.1 °F and 10.0 °F.

To set the value of the signal in the Vega chart, we call the function `vw_shiny_set_signal()`. In the context of this app:

```r
# this sets the bin-width signal in the chart
vw_shiny_set_signal("chart", name = "bin_width", value = rct_bin_width_in())
```

- `outputId`, the histogram has the outputId `"chart"`
- `name`, the Vega signal we want to set has the name `"bin_width"`
- `value`, we want to set the Vega signal to this value

Note that we are supplying a reactive expression to the `value` argument. Internally, `vw_shiny_set_signal()` observes this value, then changes the signal in the Vega chart in response to changes in `value`. Accordingly, you should think of `vw_shiny_set_signal()` as equivalent to the `shiny::observeEvent()` function.

Getting the value of the signal in a Vega chart is very similar to getting the information from an event, as described above.

The operative function in the `server()` function is `vw_shiny_get_signal()`. Here it is in the context of the app:

```r
# the signal returns the bin-width from the chart
rct_bin_width_out <- 
  vw_shiny_get_signal("chart", name = "bin_width", body_value = "value")
```

The function is called with three arguments:

- `outputId`, the scatterplot has the outputId `"chart"`
- `name`, we want to access the signal named `"bin_width"`
- `body_value`, like with the events, this is a little-bit involved...

To get a signal from Vega we add a signal listener. This includes a signal-handler, a JavaScript function that Vega runs whenever the value of the signal changes. This is very similar to an event-handler. 

For a signal-handler, the arguments are the `name` of the signal and the `value` of the signal. To make things simpler, the `vw_shiny_get_signal()` function asks for the body of a (JavaScript) function that returns the value that you want sent back to Shiny. You need only worry about returning a *value*, `vw_shiny_get_signal()` takes care of sending the *value* back to Shiny.

To make things even simpler, `body_value` can also be the name of a signal-handler in this package's handler-library. You can see all the available signal-handlers by calling `vw_handler_signal()` without arguments.

Finally, the `vw_shiny_get_signal()` function returns a reactive expression that evaluates to the latest *value* returned from the signal-handler.

To see the same logic implemented in pure JavaScript, see the [signals demo](javascript.html#demo-signals) of the JavaScript article.

## Setting and getting data

As with events and signals, you can [change datasets](https://vega.github.io/vega/docs/api/view/#view_change) in a Vega chart. In a future version of this package, you will also be able to retrieve datasets from a Vega chart.

Again, we will follow the example used in the [data section](javascript.html#signals) of the JavaScript article. Let's consider an example with a dataset that has a single observation, a point constrained to the unit-circle. We use a Shiny input to change then angle of the observation. 

To run the demonstration:

```r
vw_shiny_demo("data-set-get")
```

To be able to access a dataset in Vega(-Lite), it has to be **named** in the specification. For more information on naming datasets, please see the [Vega](https://vega.github.io/vega/docs/data) and [Vega-Lite](https://vega.github.io/vega-lite/docs/data.html#named) documentation on datasets.

In our `server()` function, we have a reactive expression, `rct_data()`, which takes the value of the slider-input and transforms it into a data frame containing a set of coordinates on the unit-circle.

To change the data in the Vega chart, we call the function `vw_shiny_set_data()`. In the context of this app:

```r
# whenever rct_data() changes, the chart will be updated
vw_shiny_set_data("chart", name = "source", value = rct_data())
```

- `outputId`, the chart has the outputId `"chart"`
- `name`, the Vega dataset we want to change has the name `"source"`
- `value`, we want to set the Vega dataset to this value

Note that we are supplying a reactive expression to the `value` argument. Internally, `vw_shiny_set_data()` observes this value, then changes the dataset in the Vega chart in response to changes in `value`. Accordingly, you should think of `vw_shiny_set_data()` as equivalent to the `shiny::observeEvent()` function.

To get a reactive copy of a named dataset from Vega, we can use the `vw_shiny_get_data()` function. When that named dataset changes in Vega, it will update the reactive returned by this function. In the context of this app:

```r
# returns the dataset in `"chart"` named `"source"`
rct_data_out <- vw_shiny_get_data("chart", name = "source")
```

To see the same logic implemented in pure JavaScript, see the [data demo](javascript.html#demo-data) of the JavaScript article.
